\vspace*{1cm}
\begin{center}
\begin{minipage}{0.8\textwidth}

\setlength\parindent{12pt}

\begin{center}
\textit{\LARGE Acknowledgment}
\end{center} \vspace{0.5cm}

The research included in this dissertation could not have been performed if not for the assistance, patience, and support of many individuals. I would like to extend my gratitude first and foremost to my thesis advisor Hans-Ulrich Heiß for the continuous support of my doctoral study and related research, for his patience, motivation, and immense knowledge. His guidance helped me during all the time of research and writing of this thesis. 

Furthermore, I like to thank Gero M\"{u}hl for motivating me to join the research group and pick up my doctoral study and for the many inspiring and insightful discussions also beyond the scope of operating systems research. I would also like to thank C\'{e}sar De Rose for his insightful comments and encouragement in order to finish my thesis and broaden my research from various perspectives.

My sincere thanks also go to Reinhardt Karnapke, Helge Parzyjegla, Matthias Diener, and Jan Richling for their support in both the research and especially the revision process that has lead to this document. Their knowledge and understanding allowed me to fully express the concepts behind this research.

I thank my fellow colleagues Daniel Graff, J\"{o}rg Schneider, Arnd Schr\"{o}ter, Jan Sch\"{o}nherr, Mohannad Nabelsee, and Stefan Sydow for the stimulating discussions, for the sleepless nights we were working together before deadlines, and for all the fun we have had in the recent years. I would also like to extend my appreciation to Gabriele Wenzel, Daniela kleine Burhoff, and Matthias Druve for their administrative support and helping me to keep my mind focused on my research.

I would additionally like to thank Anton Eisenbraun, Jens Krieg, René Sechting, and Daniel R\"{o}hrig who helped to implement the prototype and assisted me in bringing my ideas and concepts to reality.

Finally, I would like to extend my deepest gratitude to my mother Lilli Busse: Without her love, support, and understanding I could never have completed this doctoral degree.

\setlength\parindent{0pt}

\end{minipage}
\end{center}